Taking Linear Logic Apart

### 2018-04-28 doc/tlla2018/main.tex
Minor changes

\documentclass[draft,submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{example}{Example}
\input{preamble-typing}
\usepackage{more-space}
\usepackage[sectionbib,square,numbers,sort&compress]{natbib}
\renewcommand*{\eg}{e.g.\@\xspace}
\renewcommand*{\ie}{i.e.\@\xspace}

@@

In this paper, we introduce the Hypersequent Classical Processes (\hcp), which
addresses this mismatch.
The key insight is to register parallelism in the typing judgements using
hypersequents, a technique from logic which generalises judgements from one
sequent to many.
This allows us to take apart the term constructs used in Classical Processes
(\cp) to more closely match those of the standard \textpi-calculus.
We proceed as follows.
We start by introducing \cp (\cref{sec:cp}).
Then, we introduce \hcp and prove it enjoys subject reduction and progress
(\cref{sec:hcp}).
We prove several properties relating the two.
Specifically, we prove that every \cp process is an \hcp process, that every
\hcp process corresponds to several entangled \cp processes, and that \hcp
supports the same communication protocols as \cp, and no more
(\cref{sec:cp2hcp}).
Finally, we discuss related work (\cref{sec:related-work}).

@@

\begin{figure}

@@

\paragraph{Types}
Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}\noindent
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \setlength{\arraycolsep}{3pt}
    \begin{array}{lclclclclclclcl}
               \ty{(A \tens B)^\bot} & = & \ty{A^\bot \parr B^\bot}
      &\qquad& \ty{\one^\bot}        & = & \ty{\bot}
      &\qquad& \ty{(A \parr B)^\bot} & = & \ty{A^\bot \tens B^\bot}
      &\qquad& \ty{\bot^\bot}        & = & \ty{\one}
      \\       \ty{(A \plus B)^\bot} & = & \ty{A^\bot \with B^\bot}
      &\qquad& \ty{\nil^\bot}        & = & \ty{\top}
      &\qquad& \ty{(A \with B)^\bot} & = & \ty{A^\bot \plus B^\bot}
      &\qquad& \ty{\top^\bot}        & = & \ty{\nil}
    \end{array}
  \]
\end{definition}\noindent
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
\end{definition}\noindent
\begin{figure}
  \begin{center}\cpInfAx \cpInfCut \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfTens \cpInfParr \cpInfOne \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfBot \cpInfPlus1 \cpInfPlus2\end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfWith \cpInfNil \cpInfTop \end{center}
  \caption{Classical Processes, typing.}
  \label{fig:cp}
\end{figure}\noindent
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements are derived using rules in \cref{fig:cp}.
\end{definition}\noindent

@@

\paragraph{Types}
We use the same definitions for types and environments for \hcp as we used for \cp.
However, on top of this, we add the definition of a hyper-environment.
Hyper-environments are either empty, or consist of a series of typing
environments, separated by vertical bars.
A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ can be used to
type a series of $n$ entangled, but independent processes.
\begin{definition}[Hyper-environment]\label{def:hcp-hyper-environment}
  \(
  \ty{\mathcal{G}}, \ty{\mathcal{H}} :=
  \ty{\emptyhypercontext}
  \; \mid \;
  \ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}
  \)
\end{definition}\noindent
Typing judgements in \hcp associate processes with hyper-environments.
\textsc{H-Mix} composes two processes in parallel, but remembers that they are
independent on in the sequent.
\textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of
at least two unconnected processes, and connects them, eliminating the vertical
bar.
Note that each logical rule has the side condition that $\tm{x} \not\in
\mathcal{G}$, which can be read as ``you cannot act on one end-point of $x$ if
you are also holding its other end-point''.
This prevents self-locking processes, \eg
$\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes
  that the process $\tm{P}$ consists of $n$ independent, but potentially
  entangled processes, each of which communicates according to its own
  protocol $\Gamma_i$.
  Typing judgements can be constructed using the inference rules in \cref{fig:hcp}.
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.
\end{definition}\noindent
\begin{figure}
  \begin{center}\hccpInfAx \hccpInfCycle\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfMix \hccpInfHalt\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBoundTens \hccpInfParr \hccpInfOne\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBot \hccpInfPlus1 \hccpInfPlus2\end{center}
  \vspace*{0.5\baselineskip}
  \begin{center}\hccpInfWith \hccpInfNil \hccpInfTop\end{center}
  \caption{Hypersequent Classical Processes, typing.}
  \label{fig:hcp}
\end{figure}

@@

@@

It would be wonderful if each \hcp process could be translated back into a
series of \cp processes.
However, \hcp exhibits some behaviours which are impossible to directly
translate back into \cp.
For instance, in the process below, the choice sent on $\tm{x}$ will affect the
choice between $\tm{P}$ and $\tm{P'}$, even though neither has access to the
channel $\tm{x}$.
\begin{example}
  \(
  \seq[\cpCase{x}{(\piPar{P}{Q})}{(\piPar{P'}{Q'})}]{
    \Gamma, \Delta, \tmty{x}{A \with B}}
  \)
\end{example}\noindent
Instead, we will define a new relation to characterise the relation between \hcp
and \cp processes.
\begin{definition}\label{def:hcp-gk}
  The relation $\tm{P}\equiv_{\gamma\kappa}\tm{Q}$ is the congruence closure of
  the $\gamma$- and $\kappa$-rules for \hcp.
\end{definition}\noindent
This relation allows us to apply $\kappa$-rules under arbitrary prefixes.
We can show that, if we allow ourselves to do this, we can rewrite any process
to a form in which all mixes are either attached to their respective cuts or
tensors, or at the top-level.
\begin{lemma}[Disentangle]\label{lem:hcp-disentangle}
  If $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ in \hcp,
  then there exist $\seq[P_1]{\Gamma_1}, \dots, \seq[P_n]{\Gamma_n}$ in \cp,
  such that
  $\tm{P}\equiv_{\gamma\kappa}\tm{(\mtf{P_1}\ppar\dots\ppar\mtf{P_n})}$.
\end{lemma}\vspace*{-0.75\baselineskip}%
\begin{proof}
  We repeatedly apply the $\kappa{\ppar}$-rules to move the mixes upwards.
  There are three cases:
  \begin{enumerate*}[label={\alph*)}]
  \item
    if a mix gets stuck under a cut, it forms a \cp cut;
  \item
    if a mix gets stuck under a $(\tens)$, it forms a \cp $(\tens)$;
  \item
    otherwise, it moves all the way to the top.
  \end{enumerate*}
  Since the only way to continue after a $(\one)$ is by using a mix, all
  applications of $(\one)$ are now followed by an application of
  \textsc{H-Halt}, forming a \cp $(\one)$.
\end{proof}\noindent

>>>>

\documentclass[draft,submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{example}{Example}
\input{preamble-typing}
\usepackage[sectionbib,square,numbers,sort&compress]{natbib}
\renewcommand*{\eg}{e.g.\@\xspace}
\renewcommand*{\ie}{i.e.\@\xspace}
\usepackage{more-space}
\usepackage{setspace}
\renewcommand{\baselinestretch}{0.925}

@@

In this paper, we introduce the Hypersequent Classical Processes (\hcp), which
addresses this mismatch.
The key insight is to register parallelism in the typing judgements using
hypersequents~\cite{avron1991}, a technique from logic which generalises
judgements from one sequent to many.
This allows us to take apart the term constructs used in Classical Processes
(\cp) to more closely match those of the standard \textpi-calculus.
We proceed as follows.
We start by introducing \cp (\cref{sec:cp}).
Then, we introduce \hcp and prove it enjoys subject reduction and progress
(\cref{sec:hcp}).
We prove several properties relating the two.
Specifically, we prove that every \cp process is an \hcp process, that every
\hcp process corresponds to several entangled \cp processes, and that \hcp
supports the same communication protocols as \cp, and no more
(\cref{sec:cp2hcp}).
Finally, we discuss related work (\cref{sec:related-work}).

@@

\begin{figure}[htb]

@@

\end{definition}

\paragraph{Types}
Channels in \cp are typed using a session type system which corresponds to
classical linear logic.
A process communicating over a channel of type \ty{A \tens B} must be ready to
split into two independent processes (two processes which share no
channels), ready to communicate over a channel of type \ty{A} and \ty{B},
respectively, and process communicating over a channel of type \ty{A \parr B} is
dual to this, and therefore must be ready to start communicating concurrently
with two independent processes.
A process communicating over a channel of type \ty{A \plus B} can choose to
communicate either as \ty{A} or \ty{B}, and a process communicating over a
channel of type \ty{A \with B} must be able to communicate either as \ty{A} or
\ty{B}.
\begin{definition}[Types]\label{def:cp-types}
  \[
    \begin{array}{rclrcl}
      \ty{A}, \ty{B}, \ty{C}
           :=& \ty{A \tens B} &\text{pair of independent processes}
      &  \mid& \ty{\one}      &\text{unit for} \; {\tens}
      \\ \mid& \ty{A \parr B} &\text{pair of interdependent processes}
      &  \mid& \ty{\bot}      &\text{unit for} \; {\parr}
      \\ \mid& \ty{A \plus B} &\text{internal choice}
      &  \mid& \ty{\nil}      &\text{unit for} \; {\plus}
      \\ \mid& \ty{A \with B} &\text{external choice}
      &  \mid& \ty{\top}      &\text{unit for} \; {\with}
    \end{array}
  \]
\end{definition}\noindent
Duality plays a crucial role in both linear logic and session types.
In \cp, the two endpoints of a channel are assigned dual types.
This ensures that, for instance, whenever a process \emph{sends} across a
channel, the process on the other end of that channel is waiting to
\emph{receive}.
Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$.
Duality is an involutive function.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \setlength{\arraycolsep}{3pt}
    \begin{array}{lclclclclclclcl}
               \ty{(A \tens B)^\bot} & = & \ty{A^\bot \parr B^\bot}
      &\qquad& \ty{\one^\bot}        & = & \ty{\bot}
      &\qquad& \ty{(A \parr B)^\bot} & = & \ty{A^\bot \tens B^\bot}
      &\qquad& \ty{\bot^\bot}        & = & \ty{\one}
      \\       \ty{(A \plus B)^\bot} & = & \ty{A^\bot \with B^\bot}
      &\qquad& \ty{\nil^\bot}        & = & \ty{\top}
      &\qquad& \ty{(A \with B)^\bot} & = & \ty{A^\bot \plus B^\bot}
      &\qquad& \ty{\top^\bot}        & = & \ty{\nil}
    \end{array}
  \]
\end{definition}\noindent
Environments associate channels with types.
Names in environments must be unique, and environments $\ty{\Gamma}$ and
$\ty{\Delta}$ can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if
$\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} := \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}
  \) 
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\cpInfAx \cpInfCut \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfTens \cpInfParr \cpInfOne \end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfBot \cpInfPlus1 \cpInfPlus2\end{center}
  \vspace*{-.5\baselineskip}
  \begin{center}\cpInfWith \cpInfNil \cpInfTop \end{center}
  \caption{Classical Processes, typing.}
  \label{fig:cp}
\end{figure}\noindent
Typing judgements associate processes with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$ denotes
  that the process \tm{P} communicates along channels $\tm{x_1}\dots\tm{x_n}$
  following protocols $\ty{A_1}\dots\ty{A_n}$.
  Typing judgements are derived using rules in \cref{fig:cp}.
\end{definition}\noindent

@@

\paragraph{Types}
We use the same definitions for types and environments for \hcp as we used for \cp.
However, we introduce a new layer on top of sequents: hypersequents.
As \cp is a one-sided logic, and it uses the left-hand side of the
turnstile to write the process, the traditional hypersequent notation can look
confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim
that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$.
What are all the other $\Gamma$s doing there? Are they typing empty processes?
Therefore, we opt to leave out the repeated turnstile, and instead work with the
notion of ``hyper-environments''. However, we will still refer to our system as a
hypersequent system.
Hyper-environments are either empty, or consist of a series of typing
environments, separated by vertical bars.
A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of
$n$ entangled, but independent processes.
\begin{definition}[Hyper-environment]\label{def:hcp-hyper-environment}
  \(
  \ty{\mathcal{G}}, \ty{\mathcal{H}} :=
  \ty{\emptyhypercontext}
  \; \mid \;
  \ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}
  \)
\end{definition}\noindent
Typing judgements in \hcp associate processes with hyper-environments.
\textsc{H-Mix} composes two processes in parallel, but remembers that they are
independent on in the sequent.
\textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of
at least two unconnected processes, and connects them, eliminating the vertical
bar.
Note that each logical rule has the side condition that $\tm{x} \not\in
\mathcal{G}$, which can be read as ``you cannot act on one end-point of $x$ if
you are also holding its other end-point''.
This prevents self-locking processes, \eg
$\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes
  that the process $\tm{P}$ consists of $n$ independent, but potentially
  entangled processes, each of which communicates according to its own
  protocol $\Gamma_i$.
  Typing judgements can be constructed using the inference rules in \cref{fig:hcp}.
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\hccpInfAx \hccpInfCycle\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfMix \hccpInfHalt\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBoundTens \hccpInfParr \hccpInfOne\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBot \hccpInfPlus1 \hccpInfPlus2\end{center}
  \vspace*{0.5\baselineskip}
  \begin{center}\hccpInfWith \hccpInfNil \hccpInfTop\end{center}
  \caption{Hypersequent Classical Processes, typing.}
  \label{fig:hcp}
\end{figure}

@@

  \vspace*{-\baselineskip}%

@@

It would be wonderful if each \hcp process could be translated back into a
series of \cp processes.
However, \hcp exhibits some behaviours which are impossible to directly
translate back into \cp.
For instance, in the process below, the choice sent on $\tm{x}$ will affect the
choice between $\tm{P}$ and $\tm{P'}$, even though neither has access to the
channel $\tm{x}$.
\begin{example}
  \(
  \seq[\cpCase{x}{(\piPar{P}{Q})}{(\piPar{P'}{Q''})}]{
    \Gamma, \Delta, \tmty{x}{A \with B}}
  \)
\end{example}\noindent
Instead, we will define a new relation to characterise the relation between \hcp
and \cp processes.
\begin{definition}\label{def:hcp-gk}
  The relation $\tm{P}\equiv_{\gamma\kappa}\tm{Q}$ is the congruence closure of
  the $\gamma$- and $\kappa$-rules for \hcp.
\end{definition}\noindent
This relation allows us to apply $\kappa$-rules under arbitrary prefixes.
We can show that, if we allow ourselves to do this, we can rewrite any process
to a form in which all mixes are either attached to their respective cuts or
tensors, or at the top-level.
\begin{lemma}[Disentangle]\label{lem:hcp-disentangle}
  If $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ in \hcp,
  then there exist $\seq[P_1]{\Gamma_1}, \dots, \seq[P_n]{\Gamma_n}$ in \cp,
  such that
  $\tm{P}\equiv_{\gamma\kappa}\tm{(\mtf{P_1}\ppar\dots\ppar\mtf{P_n})}$.
\end{lemma}\vspace*{-0.75\baselineskip}%
\begin{proof}
  We repeatedly apply the $\kappa{\ppar}$-rules to move the mixes upwards.
  There are three cases:
  \begin{enumerate*}[label={\alph*)}]
  \item
    if a mix gets stuck under a cut, it forms a \cp cut;
  \item
    if a mix gets stuck under a $(\tens)$, it forms a \cp $(\tens)$;
  \item
    otherwise, it moves all the way to the top.
  \end{enumerate*}
  Since the only way to continue after a $(\one)$ is by using a mix, all
  applications of $(\one)$ are now followed by an application of
  \textsc{H-Halt}, forming a \cp $(\one)$.
\end{proof}\noindent
